Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2024-11-02 analysis/paper/article.qmd
writing discussion and conclusion

The academic job market is a source of uncertainty, both for job seekers, as they look for ads that are well-matched to their niche of skills and experience that have taken years to cultivate and refine, and for hiring committees as they sift through sometimes hundreds of applicants in the hope of finding a candidate that meets all of their needs. For many academics there is a perception that the last two to three decades have seen the job market become increasingly difficult for job seekers. In many fields it is common to hear that there are fewer jobs available relative to the number of PhD graduates, higher numbers of short-term temporary positions relative to secure tenure-track positions, and increasingly complex application requirements with specific statements (e.g. teaching, research, and diversity) that need to be tailed for each application, making the process of applying for jobs a full-time job of its own.

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echos findings from other fields, for example in biomedical sciences where there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from 2021–22 Faculty indicate that more than 60% of faculty positions in US universities were held by non-tenure track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, little or nothing in the way of professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities they work at [@platzerAcademicPrecarityAmerican2018].

@@

One way that some hiring committees are tackling these biases towards prestigious institutions is to provided detailed instructions to applicants on how to prepare their application materials so that the hiring committee can evaluate and rank the specific accomplishments of candidate across a variety of categories, rather than ranking based on prestige of the candidates' graduate schools. This has resulted in job ads that are highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements about their teaching and research record and plans. In a comparison of job ads from 1999–2000 and 2019–2020 published in _Anthropology News_ @HelloTristesTropes2021 noticed a trend toward increasingly asking for more materials from applicants. They found that twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported two job ads with extreme requirements, requesting nine and thirteen documents from applicants. A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, inclusion. This requirement has been much-debated due to an experiment during 2016-2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about merits and risks of requiring and using diversity statements, drawing attention to the challenges of hiring faculty from underrepresented minorities, and resulting in some prestigious universities no longer requiring diversity statements for job applicants [@guidenAreDiversityStatements].

<!--

cite these

@@

Landscape archaeology, although dominant has fluctuated over the years and has been on a downtrend since 2018-2019. Computational and digital archaeology is the second most represented method, showing an overall increasing trend, particularly since 2020-2021. Archaeobotany shows a strong cyclical trend, with a rise and fall and rise again over our study period. Archaeometry and Geoarchaeology have maintained a relatively low but steady presence in job ads, peaking in 2017-2018 and 2018-2019 and declining thereafter. Lithic analysis and Zooarchaeology are also mentioned relatively infrequently in job ads and show an inverse correlation with each other after 2018-2019.

@@

Our results reveal some of these shifts, providing insights into the dynamics of topical, methods, and geographic foci over time. While these dynamics are deeply felt by people actively seeking faculty jobs, we believe this is the first time they have been quantified at such a large scale. Trends in job ads reflect broader shifts in intellectual and practical priorities about archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe explained by a variety of processes. The downward trend in tenure-track positions during 2013-2019 related to decline in undergraduate enrollment in anthropology fields since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus.

In addition to a dip in job ads, our results show that 2019-2021 was a major inflection point in the popularity of topics and regions in job ads. Indigenous and historical archaeology and the Americas strongly increase in popularity at this time, while archaeological science, complex societies, the Mediterranean and Near East and show declines. Similarly, the number of job ads with a geographic focus on the Americas and Africa peaks during this time. These shifts in the topics and geographic foci of job ads were likely been influenced by broader cultural movements such as Black Lives Matter, protests about racial injustice, and efforts to amplify Indigenous voices. COVID-19 negatively impacted Black, American Indian and Hispanic communities with significantly higher infection and morbidity rates, drawing attention to racial and socio-economic inequality in the US. The Black Lives Matter movement, dating back to 2013, intersected profoundly with the pandemic with the murder of George Floyd by a police officer in May 2020, three months after the World Health Organization declared COVID-19 to be a pandemic. Mass protests objecting to Floyd's murder generated widespread concern about racial inequities and stimulated a broad interest in addressing systemic racial injustice. Our data suggest that US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act), universities tailored the content of the job ads to focus on topics where Black and Indigenous researchers are most numerous, ensuring that these researchers would be well-represented in the pool of applicants.

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. We found a long-term decline in topics such as the archaeology of complex societies and archaeological science, Mesoamerica and South America as a geographic focus, and methods relating to landscape archaeology. These trends are harder to explain as we cannot link their start to a discrete moment like COVID-19. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism. This might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical methods for compositional analysis. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past, increased interest in climate change, environmental sustainability, and resilience in the past may have shifted attention away from the study of monumental architecture, elite societies, political hierarchies and state systems. Other relevant factors may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation, and an increasing emphasis on more job-oriented degrees that prepare students for working in cultural heritage management rather than advanced research.

Our data on the requirements for applicants shows that application complexity has gradually increased over time. This trend is especially pronounced for Assistant Professor positions which make more demands on applicants than Associate and Full Professor positions. We saw a growing demand for research and teaching statements. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which have declined in job ads as the pandemic moves into the past. The decline in recent years for names of recommenders in the initial application may be a response to criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all.

@@

>>>>

The academic job market is a source of uncertainty, especially for job seekers, as they look for ads that are well-matched to their niche of skills and experience that have taken years to cultivate and refine. Hiring committees also face uncertainty as they strive to equitably sift through sometimes hundreds of applicants in the hope of finding a candidate that meets all of their needs and helps to realise their visions of their program and the discipline. For many academics there is a perception that the last two to three decades have seen the job market become increasingly difficult for job seekers. In many fields it is common to hear that there are fewer jobs available relative to the number of PhD graduates, higher numbers of short-term temporary positions relative to secure tenure-track positions, and increasingly complex application requirements with specific statements (e.g. teaching, research, and diversity) that need to be tailed for each application, making the process of applying for jobs a full-time job of its own.

In anthropology in the US the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. Among biological anthropologists, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echos findings from other fields, for example in biomedical sciences where there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, little or nothing in the way of professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities they work at [@platzerAcademicPrecarityAmerican2018].

@@

One way that some hiring committees are tackling these biases towards prestigious institutions is by providing detailed instructions to applicants on how to prepare their application materials so that the hiring committee can focus on evaluating and ranking the specific accomplishments of candidates across common categories, rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements about their teaching and research record and plans. In a comparison of job ads from 1999–2000 and 2019–2020 published in _Anthropology News_ @HelloTristesTropes2021 noticed a trend toward increasingly asking for more materials from applicants. They found that twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported two job ads with particularly extreme requirements, requesting nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, inclusion. This requirement has been much-debated largely due to an experiment in 2016-2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted a variety of approaches to evaluating diversity, including some universities dropping requirements for diversity statements in job applications [@guidenAreDiversityStatements].

@@

Landscape archaeology, although dominant, has fluctuated over the years and has been on a downtrend since 2018-2019. Computational and digital archaeology is the second most represented method, showing an overall increasing trend, particularly since 2020-2021. Archaeobotany shows a strong cyclical trend, with a rise and fall and rise again over our study period. Archaeometry and Geoarchaeology have maintained a relatively low but steady presence in job ads, peaking in 2017-2018 and 2018-2019 and declining thereafter. Lithic analysis and Zooarchaeology are also mentioned relatively infrequently in job ads and show an inverse correlation with each other after 2018-2019.

@@

Our results reveal some of these shifts, providing insights into the dynamics of topical, methods, and geographic foci over time. While these dynamics are deeply felt by people actively seeking faculty jobs, we believe this is the first time they have been quantified at such a large scale. Trends in job ads reflect broader shifts in intellectual and practical priorities about archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe explained by a variety of processes. Overall we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-14. This is at odds with prior work finding that contingent positions make up more than half of faculty positions. This discrepancy may be due to advertising for short-term positions typically having much more limited circulation than advertising for tenure-track jobs, and thus less likely to be added to the Academic Jobs Wiki. The downward trend in tenure-track positions during 2013-2019 maybe related to decline in undergraduate enrollment in anthropology fields since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus.

In addition to a dip in job ads, our results show that 2019-2021 was a major inflection point in the popularity of topics and regions in job ads. Indigenous and historical archaeology and the Americas strongly increase in popularity at this time, while archaeological science, complex societies, the Mediterranean and Near East and show declines. Similarly, the number of job ads with a geographic focus on the Americas and Africa peaks during this time. These shifts in the topics and geographic foci of job ads were likely been influenced by broader cultural movements such as Black Lives Matter, protests about racial injustice, and efforts to amplify Indigenous voices. COVID-19 negatively impacted Black, American Indian and Hispanic communities with significantly higher infection and morbidity rates, drawing attention to racial and socio-economic inequality in the US. The Black Lives Matter movement, dating back to 2013, intersected profoundly with the pandemic with the murder of George Floyd by a police officer in May 2020, three months after the World Health Organization declared COVID-19 to be a pandemic. Mass protests objecting to Floyd's murder generated widespread concern about racial inequities and stimulated a broad interest in addressing systemic racial injustice. Our data suggest that US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act), universities appear to have tailored the content of the job ads to focus on topics where they expect Black and Indigenous researchers to be most numerous, ensuring that these researchers would be well-represented in the pool of applicants.

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. We found a long-term decline in topics such as the archaeology of complex societies and archaeological science, Mesoamerica and South America as a geographic focus, and methods relating to landscape archaeology. These trends are harder to explain as we cannot link their start to a discrete moment like COVID-19. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism. This might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical methods for compositional analysis. Other relevant factors may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation, and an increasing emphasis on more job-oriented degrees that prepare students for working in cultural heritage management rather than advanced research. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past, increased interest in climate change, environmental sustainability, and resilience which may have shifted attention away from the study of monumental architecture, elite societies, political hierarchies and state systems.

Our data on the requirements for applicants shows that application complexity has gradually increased over time. This trend is especially pronounced for Assistant Professor positions, which make more demands on applicants than Associate and Full Professor positions. We saw a growing demand for research and teaching statements, consistent with previous work. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which may have declined in job ads as the pandemic moves into the past, in addition to the debates surrounding the experiment with diversity statements in hiring at some UC campuses. The decline in recent years for names of recommenders in the initial application may be a response to criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all.

@@

Our results show that job ads are a highly nuanced record of disciplinary history, showing shifting trends in topics, methods, and geographic foci, with higher chronological resolution than many other sources, such as undergraduate textbooks [@lyman2010american]. The short periods of time spanning the rise and fall of some of the foci we observed has implications for prospective graduate students. An area that is growing in popularity as a student begins their PhD may have peaked and be in decline before they graduate.

Jobs ads also demonstrate how archaeologists draw on current events and politics to prioritize topics, regions and methods in their hiring plans. This is consistent with prior findings, such @wilk1985ancient's analysis of publications that found trends in explanations of Mayan collapse during the 1960s-70 that correspond with major contemporary American political and social issues. This responsiveness to current events reveals that many archaeologists accept that our narratives of the past are entangled in the social and political milieu in which we work. Furthermore, it shows an active effort by archaeologists to take control of interpreting the past.
